MLH1 Results Outline

### 2020-02-11 doc/Results.rmd
reordering sections to match the outline, cutting/summarizing and simplifying

# Average MLH1 Count Model

@@

Setting up the data set for using the mixed model (base data set, HQ data set)

- only Mus musculus strains with sex matched observations

- Male gwRR have been adjusted for the PAR

- Q5 excluded

- All the effects are un-ordered factors

- male ages within 5 - 20 weeks

- Molossinus is coded as a subsp level

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE, eval=FALSE}
# MOVE this into BIVDATA SETUP!!!

#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

<!--Can I indicate the average number of cells with go into each mouse.? -->

sex: 2 female male

For now I will focus on 3 different models for comparing the subsp, sex and strain effects across relevant dependent variables.

@@

@@

## MLH1 count M1

$$mouse .mean CO ~=~ subsp * sex  + rand(1|strain) + \varepsilon$$

```{r MLH1co.av.M1, echo=FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

MLH1.MM.M1 <- lmer(mean_co ~ subsp * sex + (1|strain), data=DF.HetC.MixedModel.HQ) 
#summary(Reduced.strain.HQ)

#summary(MLH1.MM.M1)
drop1(MLH1.MM.M1, test="Chisq") # p = 0.0002045 ***

as.data.frame(round(summary(MLH1.MM.M1)$coefficients, 6))

#only the interaction term shows up, when the 2 terms are additive, they both show up in the drop1 results
anova(MLH1.MM.M1)

exactRLRT(MLH1.MM.M1) #< 2.2e-16  Random strain effect, the variance due to random strain effect is NOT 0

Above is the mixed model M1,

## MLH1 count M2

```{r MixedModel.2, echo=FALSE,  digits = 3}
#strain = fixed
lm.strain.fixed.full <- lm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

anova(lm.strain.fixed.full)

#this table of results is broken -- might be too long -- lm
#lm has different names
as.data.frame(round( summary(lm.strain.fixed.full)$coefficients, 6))

#summary(strain.fixed.full)
#when strain added as fixed, MSM, PWD and WSB are signifcant, subsp*sex is most significant

#non of the 3-way interactions are significant, very puzzeled that WSB is significant and non of ht ehigh Musc are 
```

@@

## MLH1 count M3

@@

#### Summary of Mixed models

@@

### MLH1 variance M1

@@

### MLH1 variance M2

@@

### MLH1 variance M3

@@

# Female Specific Polymorphism

I should run the same questions for the female data --

musc.female.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))
musc.male.plot

#variance
musc.female.VAR.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
musc.female.VAR.plot

Mol.female.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Mol.female.plot

Dom.female.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom feale ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)
Dom.female.plot

```{r Female.specific.test, echo=FALSE}

lm.female.strain.subsp <- lm(mean_co ~ subsp * strain, data= female_mouse.avs_for.analysis)
summry.female.strain.effects <- summary(lm.female.strain.subsp)
#strainG, 1.06e-05 ***
#strainLEW  0.0273 *

lm.female.strain <- lm(mean_co ~ strain, data= female_mouse.avs_for.analysis)
lm.female.strain_sum <- summary(lm.female.strain)
#G  1.06e-05 ***
#LEW   0.0273 *
#PWD  0.0810 .
#MSM  3.02e-05 ***

sans_G_female <- female_mouse.avs_for.analysis %>% filter(strain != "G")
mean.all.other <- mean(sans_G_female$mean_co)

# Male Specific Polymorphism

```{r Male.specific.test, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect

#1. determine which strain effects are sig

lm.male.strain.subsp <- lm(mean_co ~ subsp * strain, data= just.male_mouse.av)
summry.male.strain.effects <- summary(lm.male.strain.subsp)

lm.male.strain <- lm(mean_co ~ strain, data= just.male_mouse.av)
lm.male.strain_sum <- summary(lm.male.strain)

sans_LewRec_male <- just.male_mouse.av %>% filter(Rec.group == 0)
#mean(sans_LewRec_male$mean_co)

msm, pwd skive are when all other male strains are added. skive is less sig than the others -- but it's still sig.  Given these results I can group SKIVE, PWD and MSM into the 'high rec' group.

*PWD, MSM, SKIVE are in 'high rec group'*

# Proportion of chromosome classes

@@

```{r test.Chrm.Prop, echo=FALSE}

@@

                  nbiv = sum(tot.biv ),              
                 n0CO = sum(n0CO ),
                  n1CO = sum(n1CO ),
                 n2CO = sum(n2CO ),
                  n3CO = sum(n3CO )

>>>>

# gwRR estimates

make table of the strain averages

```{r MLH1.summ.stats, echo=FALSE}

#table for the means by strain

## Female Specific Analysis

musc.female.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = mean_co, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of MLH1 distributions")+scale_color_manual(values=colors_of_strains)+ylim(c(20,34))
musc.female.plot

#variance plots
musc.female.VAR.plot <- ggplot(Musc.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y =var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Musc female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Mol.female.plot <- ggplot(Mol.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Mol female of within mouse MLH1 distributions")+scale_color_manual(values=colors_of_strains)

Dom.female.plot <- ggplot(Dom.female_mouse.av, aes(color= strain))+geom_jitter(aes(x = as.factor(strain), y = var, color= strain, alpha=0.5), width = .2 )+ ggtitle("Dom feale ofwithin mouse  MLH1 distributions")+scale_color_manual(values=colors_of_strains)

```{r female.strain.table, echo=FALSE}

#female strain table
#female_mouse.avs_for.analysis
F.strain.table <- ddply(female_mouse.avs_for.analysis, c("strain"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  = sum(Ncells),
                        mean_co =as.numeric(format(round(mean(mean_co), 3 ), nsmall=3) ),

                        mean.mouse_cV = mean(cV, na.rm = TRUE),

                        mean.mouse_var = mean(var, na.rm = TRUE ),

                        mean.mouse_sd   = mean(sd),

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se)
    )

```{r Female.specific.glm, echo=FALSE}

glm.female.strain.subsp <- lm(mean_co ~ subsp * strain, data= female_mouse.avs_for.analysis)

summry.female.strain.effects <- summary(glm.female.strain.subsp)

#write.table(summry.female.strain.effects$coefficients, "~./MLH1repo/summry.female.strain.effects.csv", sep=",",row.names = TRUE)

lm.female.strain <- glm(mean_co ~ strain, data= female_mouse.avs_for.analysis)
lm.female.strain_sum <- summary(lm.female.strain)

#write.table(lm.female.strain_sum$coefficients, "~./MLH1repo/lm.female.strain.csv", sep=",",row.names = TRUE)

sans_high_female <- female_mouse.avs_for.analysis %>% filter(strain != "G") %>%  filter(strain != "MSM")
mean.all.other <- mean(sans_high_female$mean_co)

#how much greater is MSM female
MSM_female <-  female_mouse.avs_for.analysis %>% filter(strain == "MSM")
mean.MSN.female <- mean(MSM_female$mean_co)

sans_MSM_female <- female_mouse.avs_for.analysis %>% filter(strain != "MSM")
mean.all.other <- mean(sans_MSM_female$mean_co)

mean.MSN.female / mean.all.other

The range of strain means is `r min()`

the ranges of mouse means for each strain is X

MSM, G, LEW have significant strain effects.

## Male Specific Analysis

```{r male.strain.table, echo=FALSE}

#female strain table
#female_mouse.avs_for.analysis
M.strain.table <- ddply(just.male_mouse.av, c("strain"), summarise,
                        Nmice =length(unique(mouse)),
                        Ncells  = sum(Ncells),
                        mean_co =as.numeric(format(round(mean(mean_co), 3 ), nsmall=3) ),

                        mean.mouse_cV = mean(cV, na.rm = TRUE),

                        mean.mouse_var = mean(var, na.rm = TRUE ),

                        mean.mouse_sd   = mean(sd),

                        #Standard error (mice vs cells)
                        mean.mouse_se  = mean(se)
    )

```{r Male.specific.glm, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect
glm.male.strain.subsp <- lm(mean_co ~ subsp * strain, data= just.male_mouse.av)

summry.male.strain.effects <- summary(glm.male.strain.subsp)

#write.table(summry.male.strain.effects$coefficients , "~./MLH1repo/summry.male.strain.effects.csv", sep=",",row.names = TRUE)

#1. determine which strain effects are sig
#M2
lm.male.strain <- glm(mean_co ~ strain, data= just.male_mouse.av)
lm.male.strain_sum<- summary(lm.male.strain)

#write.table(lm.male.strain_sum$coefficients , "~./MLH1repo/lm.male.strain_sum.csv", sep=",",row.names = TRUE)

#how much bigger are the high rec male means?
sans_LewRec_male <- just.male_mouse.av %>% filter(Rec.group == 0)
#mean(sans_LewRec_male$mean_co)

the ranges of means are X

PWD, MSM and SKIVE have strain significant strain effects so they will be designedated into the high rec group.

# Analysis for Evolutionary Patterns

In order to understand the variance within an evolutionary framework we fit the mouse gwRRs to a mixed model, which specified subspecies, sex and the interaction as fixed effects. The subspecies effect - is a proxy for measuring the degree of divergence across subspecies. The strains were coded as a random effect to -simulate / approximate / the standing genetic variation across subspecies.

@@

<!--
Setting up the data set for using the mixed model (base data set, HQ data set)

- only Mus musculus strains with sex matched observations  
- Male gwRR have been adjusted for the PAR
- Q5 excluded
- All the effects are un-ordered factors  
- male ages within 5 - 20 weeks
- Molossinus is coded as a subsp level

sex: 2 female male  
-->

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE, eval=FALSE}
#DF.HetC.MixedModel.HQ

@@

@@

```{r MLH1co.MM, echo=FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

MLH1.MM.M1 <- lmer(mean_co ~ subsp * sex + (1|strain), data=DF.HetC.MixedModel.HQ) 
#summary(MLH1.MM.M1)

MM.Fixed.pval <- drop1(MLH1.MM.M1, test="Chisq")$`Pr(Chi)`[2]
MM.rand.pval <- exactRLRT(MLH1.MM.M1)$p.value

#as.data.frame(round(summary(MLH1.MM.M1)$coefficients, 6))

#only the interaction term shows up, when the 2 terms are additive, they both show up in the drop1 results

The significance of the interaction effect for the 2 fixed effects is `r MM.Fixed.pval`. The random effect for strain effect is `r MM.rand.pval`.

### glm followup

To follow up on the mixed model results, we also ran glm / linear models to get a better idea of the strain by sex interaction effects.

@@

```{r MixedModel.2, echo=FALSE,  digits = 3}
#strain = fixed
glm.strain.fixed.full <- glm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

summ <- summary(glm.strain.fixed.full)

#write.table(summ$coefficients , "~./MLH1repo/glm.strain.fixed.full.csv", sep=",",row.names = TRUE)

@@

@@

@@

@@

@@

@@

# Proportion of Chromosome Classes

In order to decompose the cell wide rate, we decided to look at the proportion or chromosomes with different numbers of COs)

@@

```{r test.Chrm.Prop, echo=FALSE, eval=FALSE}

@@

                  nbiv = sum(tot.biv ),              
                 n0CO = sum(n0CO ),
                  n1CO = sum(n1CO ),
                 n2CO = sum(n2CO ),
                  n3CO = sum(n3CO )
                          )
